Relational Parametricity for Algebraically Indexed Types

### 2012-05-22 paper/old-syntax.tex
Update to the draft paper
- Split each section into a separate file

- Rewrite the description of the general framework: There is no major
change to the framework itself, except for allowing primitive types to
have more than one index (so that we can have numbers indexed by a
scaling factor and an offset, for instance). I have rewritten most of
the explanation though, in the hope that it is more understandable
now. A major change to the presentation is that I have changed the
order of syntax and semantics to be interleaved: I define the
semantics of types (index-erasure and relational) before

\documentclass[natbib,preprint]{sigplanconf}

\usepackage[usenames]{color}
\definecolor{citationcolour}{rgb}{0,0.4,0.2}
\definecolor{linkcolour}{rgb}{0,0,0.8}
\usepackage{hyperref}
\hypersetup{colorlinks=true,
            urlcolor=linkcolour,
            linkcolor=linkcolour,
            citecolor=citationcolour,
            pdftitle=Relational Parametricity for Algebraically-Indexed Types,
            pdfauthor={Robert Atkey, Neil Ghani, Patricia Johann, Andrew Kennedy},
            pdfkeywords={}}  
\def\sectionautorefname{Section}
\def\subsectionautorefname{Section}

\title{Relational Parametricity for Algebraically Indexed Types}

\authorinfo{Robert Atkey\and Neil Ghani\\\and Patricia Johann}
           {University of Strathclyde}
           {\{Robert.Atkey,Neil.Ghani,Patricia.Johann\}@cis.strath.ac.uk}

\authorinfo{Andrew Kennedy}
           {Microsoft Research}
           {}

\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage[all]{xy}

\newcommand{\sepbar}{\mathrel|}

\newcommand{\Rel}{\mathrm{Rel}}
\newcommand{\relArrow}{\mathrel{\widehat\to}}
\newcommand{\relTimes}{\mathrel{\widehat\times}}
\newcommand{\relSum}{\mathrel{\widehat+}}

\newcommand{\setOfIntegers}{\mathbb{Z}}
\newcommand{\setOfBooleans}{\{\tmTT, \tmFF\}}

\newcommand{\id}{\mathrm{id}}

\newcommand{\indexOp}[1]{\texttt{#1}}
\newcommand{\idxTms}[2]{\mathrm{IdxTm}(#1 \vdash #2)}
\newcommand{\idxTmAS}[1]{\mathrm{IdxTm}(#1)}

\newcommand{\sortType}{\texttt{type}}
\newcommand{\tyInt}{\texttt{int}}
\newcommand{\tyBool}{\texttt{bool}}
\newcommand{\tyUnit}{\texttt{unit}}
\newcommand{\tyPrim}[2]{\texttt{#1}\langle #2 \rangle}
\newcommand{\tyPrimNm}[1]{\texttt{#1}}

\newcommand{\tyArr}{\to}
\newcommand{\tyProduct}{\times}
\newcommand{\tyX}[1]{\texttt{X}\langle #1 \rangle}
\newcommand{\isType}{\textsf{ type}}
\newcommand{\isCtxt}{\textsf{ ctxt}}
\newcommand{\Ty}{\textsf{Ty}}
\newcommand{\ty}{\textsf{ty}}
\newcommand{\subst}[2]{\mathrm{Subst}(#1,#2)}

\newcommand{\tmTT}{\texttt{tt}}
\newcommand{\tmFF}{\texttt{ff}}

\newcommand{\relEnv}[1]{\mathcal{#1}}
\newcommand{\tySem}[1]{\llbracket #1 \rrbracket^{\mathcal{T}}}
\newcommand{\ctxtSem}[1]{\llbracket #1 \rrbracket^{\mathcal{C}}}
\newcommand{\tmSem}[1]{\llbracket #1 \rrbracket^{\mathit{tm}}}
\newcommand{\tyPrimSem}[1]{\llbracket \tyPrimNm{#1} \rrbracket^{\mathcal{T}_0}}
\newcommand{\rsem}[3]{\llbracket #1 \rrbracket^{\mathcal{R}}_{#2}{#3}}
\newcommand{\extends}[2]{\mathsf{ext}(#1,#2)}

\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newcommand{\lemref}[1]{\hyperref[#1]{Lemma \ref*{#1}}}
\newcommand{\thmref}[1]{\hyperref[#1]{Theorem \ref*{#1}}}
\newcommand{\propref}[1]{\hyperref[#1]{Proposition \ref*{#1}}}
\newcommand{\corref}[1]{\hyperref[#1]{Corollary \ref*{#1}}}
\newcommand{\conref}[1]{\hyperref[#1]{Conjecture \ref*{#1}}}
\newcommand{\exref}[1]{\hyperref[#1]{Example \ref*{#1}}}

\begin{document}

\maketitle

\begin{abstract}

\end{abstract}

\category{D.1.1}{Programming techniques}{Applicative (functional)
  programming} \category{D.2.4}{Software Engineering}{Software/Program
  Verification} \category{D.3.3}{Programming Languages}{Language
  Constructs and Features---Data types and structures}

\terms
  Languages, Theory, Types

\keywords
  parametricity, dimension types, computational geometry

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

\begin{enumerate}
\item Introduction using the area-of-a-triangle example
\item Narrowing of the semantic gap by use of advanced type systems
\item Importance of the use of symmetry in computer science,
  mathematics and the physical sciences
\item Development of the theory of parametricity
\end{enumerate}

\subsection{Contributions}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background and Related Work}

Reynolds (Parametricity), Wadler (Free Theorems).

Kennedy (Units of Measure)

Something about symmetry?

@@

@@

\begin{figure*}[t]
  \centering
  Well-indexed Types
  \begin{mathpar}
    \inferrule* [right=TyPrimIdx]
    {\tyPrimNm{X} \in \mathit{PrimType}(s) \\ \Delta \vdash e : s}
    {\Delta \vdash \tyPrim{X}{e} : \sortType}

    \inferrule* [right=TyPrimNoIdx]
    {\tyPrimNm{X} \in \mathit{PrimType}(\cdot)}
    {\Delta \vdash \tyPrimNm{X} : \sortType}

    \inferrule* [right=TyUnit]
    { }
    {\Delta \vdash \tyUnit : \sortType}

    \inferrule* [right=TyArr]
    {\Delta \vdash A : \sortType \\ \Delta \vdash B : \sortType}
    {\Delta \vdash A \tyArr B : \sortType}

    \inferrule* [right=TyProd]
    {\Delta \vdash A : \sortType \\ \Delta \vdash B : \sortType}
    {\Delta \vdash A \tyProduct B : \sortType}

    \inferrule* [right=TySum]
    {\Delta \vdash A : \sortType \\ \Delta \vdash B : \sortType}
    {\Delta \vdash A + B : \sortType}

    \inferrule* [right=TyForall] %FIXME: macroize forall
    {\Delta, i : s \vdash A : \sortType}
    {\Delta \vdash \forall i \mathord: s. A : \sortType}
  \end{mathpar}

  \medskip
  Equality of well-indexed types
  \begin{mathpar}
    \inferrule* [right=TyEqPrimIdx]
    {\Delta \vdash e_1 \equiv e_2 : s}
    {\Delta \vdash \tyPrim{X}{e_1} \equiv \tyPrim{X}{e_2} : \sortType}

    \inferrule* [right=TyEqPrimNoIdx]
    { }
    {\Delta \vdash \tyPrimNm{X} \equiv \tyPrimNm{X} : \sortType}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Example Systems}\label{sec:example-systems}

\subsection{Units of Measure}\label{sec:units-example}

We now show that our general framework encompasses Kennedy's
units-of-measure system \cite{kennedy97relational}. For our indexing
algebraic theory, we have a single sort $\mathsf{unit}$ intended to
represent composite units such as $\mathit{m}\ \mathit{s}^{-1}$ or
$\mathit{kg}\ \mathit{m}^{-2}$. Units of measure form an abelian
group, so we have three operations:
\begin{mathpar}
  1 : \mathsf{unit},

  \cdot : \mathsf{unit}, \mathsf{unit} \to \mathsf{unit},

  -^{-1} : \mathsf{unit} \to \mathsf{unit},
\end{mathpar}
subject to the normal abelian group axioms. We will elide the group
multiplication operation and just write multiplication as
juxtaposition: $u_1u_2$.

We let $\mathit{PrimType}(\mathsf{unit}) = \{\tyPrimNm{num}$, and
$\mathit{PrimType}(\cdot) = \{\tyPrimNm{rational}\}$. The intended
reading is that the type $\tyPrim{num}{u_1u^2_2}$ represents
quantities in units of $u_1u^2_2$, while values of type
$\tyPrimNm{rational}$ represent dimensionless constants. We make a
distinction between $\tyPrim{num}{1}$ and $\tyPrimNm{rational}$ for
technical reasons arising from the relational interpretation below: we
want the relation on $\tyPrim{num}{1}$ to be the identity relation,
and the existence of $\mathit{toRat}$ and $\mathit{fromRat}$ forces
this\footnote{Kennedy assumes that all relational environments set the
  relation on $\tyPrim{num}{1}$ to be the identity relation, but we
  cannot do this in general, because we do not necessarily have a
  group structure in the indexing theory.}. To write programs using
the $\tyPrimNm{num}$ type, we assume an typing context
$\Gamma_{\textrm{units}}$ consisting of the following typed
operations:
\begin{eqnarray*}
  \mathit{toRat} & : & \tyPrim{num}{1} \to \tyPrimNm{rational} \\
  \mathit{fromRat} & : & \tyPrim{num}{1} \to \tyPrimNm{rational} \\
  0 & : & \forall u. \tyPrim{num}{u} \\
  1 & : & \tyPrim{num}{1} \\
  + & : & \forall u. \tyPrim{num}{u} \to \tyPrim{num}{u} \to \tyPrim{num}{u} \\
  - & : & \forall u. \tyPrim{num}{u} \to \tyPrim{num}{u} \to \tyPrim{num}{u} \\
  * & : & \forall u_1 u_2. \tyPrim{num}{u_1} \to \tyPrim{num}{u_2} \to \tyPrim{num}{u_1u_2} \\
  / & : & \forall u_1 u_2. \tyPrim{num}{u_1} \to \tyPrim{num}{u_2} \to \tyPrim{num}{u_1u_2^{-1}} + 1 \\
  < & : & \forall u. \tyPrim{num}{u} \to \tyPrim{num}{u} \to \tyBool
\end{eqnarray*}
There are constants for zero and one, but only the constant for zero
is polymorphic in the units. In \autoref{sec:units-semantics} below,
we will following Kennedy by interpreting units of measure in terms of
\emph{scalings} by rational numbers, where it will be apparent that
the constant one cannot be unit-polymorphic: we can scale $0$ and
still get $0$, but we cannot scale $1$ by an arbitrary rational number
and expect to get $1$.

\subsection{Euclidean Plane Geometry}\label{sec:2d-example}

We now consider a type system for writing programs that are invariant
under the translational symmetries of the Euclidean plane. We assume a
single sort $\mathsf{translation}$. The sort $\mathsf{translation}$ is
intended to represent translations in the plane, and comes equipped
with abelian group structure that we write additively:
\begin{mathpar}
  0 : \mathsf{translation}

  + : \mathsf{translation},\mathsf{translation} \to \mathsf{translation}

  - : \mathsf{translation} \to \mathsf{translation}
\end{mathpar}
In indexing expressions involving variables of sort
$\mathsf{translation}$, we use the shorthand $t_1 - t_2$ to stand for
$t_1 + (-t_2)$.

There are two primitive types:
$\mathit{PrimType}(\mathsf{translation}) = \{\tyPrimNm{vec}\}$, and
$\mathit{PrimType}(\cdot) = \{\tyPrimNm{real}\}$. The types
$\tyPrim{vec}{t}$ represent two-dimensional vectors, indistinguishable
up to some translation $t$. Values of type $\tyPrimNm{real}$ are just
real numbers. To write programs using these types we assume a typing
context $\Gamma_{\mathit{2D}}$ consisting of the following operations:
\begin{eqnarray*}
  + & : & \forall t_1 t_2\mathord:\mathsf{translation}.\tyPrim{vec}{t_1} \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 + t_2} \\
  - & : & \forall t\mathord:\mathsf{translation}. \tyPrim{vec}{t} \to \tyPrim{vec}{-t} \\
  0 & : & \tyPrim{vec}{0} \\
  * & : & \tyPrimNm{real} \to \tyPrim{vec}{0} \to \tyPrim{vec}{0} \\
  \mathrm{getX} & : & \tyPrim{vec}{0} \to \tyPrimNm{real} \\
  \mathrm{getY} & : & \tyPrim{vec}{0} \to \tyPrimNm{real} \\
  \mathrm{vec}  & : & \tyPrimNm{real} \to \tyPrimNm{real} \to \tyPrim{vec}{0}
\end{eqnarray*}
The first three operations express the abelian group structure of
2-dimensional real vectors. Note how the indexing tracks the
operations on vectors: e.g.~addition on vectors is reflected by
addition on the indicies. The final four operations are operations
that are not invariant under arbitrary translation: multiplication by
a scalar, and projection from and constructon of vectors. For the
purposes of examples, we also assume a standard set of arithmetic and
other operations on real numbers.

As an example of the use of the above operations, consider the
following type of a function for computing the area of a triangle
described by three points. Since the area of a triangle is invariant
under translations that are applied to all three points, we can
quantify over all translations $t$:
\begin{displaymath}
  \mathit{area} : \forall t\mathord:\mathsf{translation}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrimNm{real}
\end{displaymath}
Since the result is invariant under translation, it safe to assume
that one of the vertexes of the triangle is the origin, and only
specify two points to describe the triangle. So we might expect that
the type of $\mathit{area}$ is isomorphic to this type:
\begin{displaymath}
  \mathit{area'} : \tyPrim{vec}{0} \to \tyPrim{vec}{0} \to \tyPrimNm{real}
\end{displaymath}
Given $\mathit{area'}$, we can define $\mathit{area}$ using the
operations in $\Gamma_{\mathit{2D}}$:
\begin{displaymath}
  \mathit{area} = \Lambda t. \lambda v_1 v_2 v_3.\ \mathit{area'}\ (v_2 - v_1)\ (v_3 - v_1)
\end{displaymath}
and vice versa, we can define $\mathit{area'}$ in terms of
$\mathit{area}$:
\begin{displaymath}
  \mathit{area'} = \lambda v_1 v_2. \mathit{area}\ [0]\ 0\ v_1\ v_2
\end{displaymath}
In \autoref{sec:2d-semantics} below, we will show that these two
translations form a type isomorphism between the types of
$\mathit{area}$ and $\mathit{area'}$.

\paragraph{Alternative Transformation Groups?}
We should also consider other groups of transformations for indexing:
isometries, orthogonal transformations, rotations and reflections.

\subsection{Security Typing and Non-interference}

This example shows how our framework can accommodate non-interference
properties common in security type systems. This example has a
different flavour to the two previous examples because it does not
concern invariance under the action of some group. We assume a single
sort $\mathsf{principal}$, intended to represent (possibly composite)
principals in a system. Principals are combined using the connectives
of boolean logic:
\begin{mathpar}
  \top : \mathsf{principal}

  \bot : \mathsf{principal}

  \land : \mathsf{principal},\mathsf{principal} \to \mathsf{principal}

  \lor : \mathsf{principal},\mathsf{principal} \to \mathsf{principal}

  \Rightarrow : \mathsf{principal},\mathsf{principal} \to \mathsf{principal}

  \lnot : \mathsf{principal} \to \mathsf{principal}
\end{mathpar}
For the equational theory of principals, we assume all the axioms of
boolean algebra.

Let $\mathit{PrimType}(\mathsf{principal}) = \{\tyPrimNm{T}\}$. The
intended reading of a type $\tyPrim{T}{p}$ for some principal $p$ is
that the existence of a value of this type indicates that the
principal $p$ is ``happy''. To write programs, we assume the following
typing context $\Gamma_{\mathit{sec}}$ of operations:
\begin{eqnarray*}
  \mathrm{proj} & : & \forall p_1 p_2\mathord:\mathsf{principal}.\ \tyPrim{T}{p_1 \land p_2} \to \tyPrim{T}{p_1} \\
  \mathrm{combine} & : & \forall p_1 p_2\mathord:\mathsf{principal}.\ \tyPrim{T}{p_1} \to \tyPrim{T}{p_2} \to \tyPrim{T}{p_1 \land p_2}
\end{eqnarray*}

To illustrate the use of this application of our framework, we define
a type synonym, for each type $A$ and principal $p$:
\begin{displaymath}
  T_pA = \tyPrim{T}{p} \to A
\end{displaymath}
For every principal $p$, we can endow the types $T_p-$ with the
structure of a monad. This is due to the fact that it is an instance
of the ``environment'' (or ``reader'') monad. Explicitly, the
$\mathit{return}$ and $>>=$ of these monads are given by:
\begin{eqnarray*}
  \mathit{return} & : & A \to T_pA \\
  \mathit{return} & = & \lambda a\ t.\ a \\
  \\
  (>>=) & : & T_pA \to (A \to T_pB) \to T_pB \\
  (>>=) & = & \lambda a\ f\ t.\ f\ (a\ t)\ t
\end{eqnarray*}
The intended reading of a value of type $T_pA$ is as a piece of data
of type $A$ that can only be accessed if the principal $p$ is
happy. The existence of such a monad should allow us to draw parallels
between our system and the Dependency Core Calculus (DCC) system of
Abadi \emph{et al.} \cite{abadi99core}. The definition of a DCC monad
as an instance of an environment monad follows Tse and Zdancewic's
interpretation of DCC in System F, where they use System F
parametricity to ensure non-interference properties.

There is also a coercion operator between instances of $T_pA$ for
different principals. Reading the type in terms of happiness, this
says that if we have a piece of data that is protected by $p$, and
happiness of $q$ implies happiness of $p$, then this data can also be
accessed by $q$.
\begin{eqnarray*}
  \mathit{coerce} & : & \tyPrim{T}{q \Rightarrow p} \to T_pA \to T_qA \\
  \mathit{coerce} & = & \lambda t_{qp}\ a\ t_q.\ a\ (\mathrm{combine}\ t_{qp}\ t_q)
\end{eqnarray*}
Here we have used the fact that the result of the application of
$\mathrm{combine}$ has type $\tyPrim{T}{(q \Rightarrow p) \land
  q}$. By the laws of boolean algebras---which we have assumed for our
indexing theory---this is equal to the type $\tyPrim{T}{p}$. Thus we
get access to the data stored in $a$, which is protected by $p$.

Now suppose we have a term $M$ with the following typing:
\begin{displaymath}
  p,q:\mathsf{principal}; t_{qp} : \tyPrim{T}{q \Rightarrow p}, x : T_q (1 + 1) \vdash M : T_p (1 + 1)
\end{displaymath}
Intuitively, given the operations $\Gamma_{\mathit{sec}}$ we have
defined above, it should be the case that $M$ can only be equivalent
to the program which is constantly $\lambda t. \mathrm{inl}(*)$ or
$\lambda t. \mathrm{inr}(*)$, because there is no useful data to be
gained from the $t_{qp}$ argument, and a value of $\tyPrim{T}{q}$
cannot be generated in order to get access to the data within the $x$
argument. The application of our framework that we have presented in
this section has a ``non-interference'' property: the value of $x$
cannot interfere with the output. In \autoref{sec:sec-semantics} we
show that this is indeed the case.

\paragraph{Alternative Logics?}
Above, we assumed that the indexing theory of principals was the
theory of boolean algebras. As a consequence, a value of type $T_{p
  \lor \lnot p}A$ can always be accessed, due to the exlcuded middle
property of boolean logic. An interesting question is: what happens we
we consider alternative logics. If we take the theory of Heyting
algebras for our theory of principals, then data typed as $T_{p \lor
  \lnot p}A$ can only be accessed if either $p$ or $\lnot p$ can be
demonstrated. Considering an alternative logical theory alters the
non-interference property of the type system.

\paragraph{Alternative Presentations?}
In the above we presented the theory of boolean logic by means of the
equational theory of boolean algebras. This isn't necessarily the most
convenient way of presenting a logic. An alternative approach would be
to consider a trivial equational theory with no axioms, and
incorporate the rules of logic consequence as operations in
$\Gamma_{\mathit{sec}}$. For example, as a Hilbert system:
\begin{eqnarray*}
  \mathit{MP} & : & \forall p,q.\ \tyPrim{T}{p \Rightarrow q} \to \tyPrim{T}{p} \to \tyPrim{T}{q} \\
  \mathit{k} & : & \forall p,q.\ \tyPrim{T}{p \Rightarrow q \Rightarrow p} \\
  \mathit{s} & : & \forall p,q,r.\ \tyPrim{T}{(p \Rightarrow q \Rightarrow r) \Rightarrow (p \Rightarrow q) \Rightarrow p \Rightarrow r}
\end{eqnarray*}
Plus axioms for the other connectives.  Or possibly some kind of
encoding of a natural deduction system?

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Semantics and Parametricity Theorem}

There is an index-erasure semantics, and a relational semantics. The
relational semantics allows us to derive free theorems about the
index-erasure semantics.

\subsection{Index Erasure Semantics}

We now define a semantics of types and terms that ignores all indexing
information. By defining this semantics we can easily see that the
indexing information can be discarded before execution.

\subsubsection{Index-Erasure Semantics of Types}

For each primitive type $\tyPrimNm{X} \in \mathit{PrimType}(s)$, for
some sort $s$, and for the primitive types $\tyPrimNm{X} \in
\mathit{PrimType}(\cdot)$, we assume that $\tyPrimNm{X}$ is assigned a
set $\tyPrimSem{X}$. We extend this to assign a set to every
well-formed type by induction on the type structure. To interpret the
universal quantifier, we simply ignore the indexing:
\begin{eqnarray*}
  \tySem{\tyUnit} & = & \{*\} \\
  \tySem{\tyPrim{X}{e}} & = & \tyPrimSem{X} \\
  \tySem{\tyPrim{X}} & = & \tyPrimSem{X} \\
  \tySem{A \tyProduct B} & = & \tySem{A} \times \tySem{B} \\
  \tySem{A \tyArr B} & = & \tySem{A} \to \tySem{B} \\
  \tySem{A + B} & = & \tySem{A} + \tySem{B} \\
  \tySem{\forall i\mathord:s. A} & = & \tySem{A}
\end{eqnarray*}

Since our definition has completely ignored the indexing, and type
equality is defined as an extension of index equality, we can easily
see that equal types have equal denotations in our index-erasure
semantics:
\begin{lemma}\label{lem:tyeq-erasure}
  If $\Delta \vdash A \equiv B : \sortType$ then $\tySem{A} =
  \tySem{B}$.
\end{lemma}

Moreover, the index-erasure semantics is insensitive to index
substitutions.
\begin{lemma}\label{lem:tysubst-erasure}
  If $\Delta \vdash A : \sortType$ and $\sigma : \Delta' \Rightarrow
  \Delta$, then $\tySem{\sigma^*A} = \tySem{A}$.
\end{lemma}

For any typing context $\Delta \vdash \Gamma \isCtxt$, we define its
index-erasure semantics $\ctxtSem{\Gamma}$ by induction on its
structure: $\ctxtSem{\epsilon} = 1$ and $\ctxtSem{\Gamma, x : A} =
\ctxtSem{\Gamma} \times \tySem{A}$, where $1$ is the canonical
one-element set.

\subsubsection{Index-Erasure Semantics of Terms}

For a well-formed term $\Delta; \Gamma \vdash e : A$, we define a
function $\tmSem{e} : \ctxtSem{\Gamma} \to \tySem{A}$, completely
ignoring the indexing information. In light of
\lemref{lem:tyeq-erasure} and \lemref{lem:tysubst-erasure}, we may
define this function directly on the syntax of well-typed terms,
rather than on the typing derivations. We define $\tmSem{e}$ by the
following clauses:
\begin{eqnarray*}
  \tmSem{x}\eta & = & \eta_x \\
  \tmSem{(M,N)}\eta & = & (\tmSem{M}\eta, \tmSem{N}\eta) \\
  \tmSem{\pi_1M}\eta & = & \pi_1(\tmSem{M}\eta) \\
  \tmSem{\pi_2M}\eta & = & \pi_2(\tmSem{M}\eta) \\
  \tmSem{\mathrm{inl}\ M}\eta & = & \mathrm{inl}\ (\tmSem{M}\eta) \\
  \tmSem{\mathrm{inr}\ M}\eta & = & \mathrm{inr}\ (\tmSem{M}\eta) \\
  \left\llbracket
    \begin{array}{l}
      \textrm{case}\ M\ \textrm{of}\\
      \textrm{inl}\ x.N_1;\\
      \textrm{inr}\ y.N_2
    \end{array}\right\rrbracket^{\mathit{tm}}\eta & = &
  \left\{
    \begin{array}{ll}
      \tmSem{N_1}(\eta,a) & \textrm{if }\tmSem{M}\eta = \mathrm{inl}(a) \\
      \tmSem{N_2}(\eta,b) & \textrm{if }\tmSem{M}\eta = \mathrm{inr}(b)
    \end{array}
  \right. \\
  \tmSem{\lambda x. M}\eta & = & \lambda v. \tmSem{M}(\eta, v) \\
  \tmSem{M N}\eta & = & (\tmSem{M}\eta) (\tmSem{N}\eta) \\
  \tmSem{\Lambda i.\ M}\eta & = & \tmSem{M}\eta \\
  \tmSem{M[e]}\eta & = & \tmSem{M}\eta
\end{eqnarray*}
In the above, we have written $\eta_x$ to denote the appropriate
projection from $\eta$ to get the position corresponding to the
variable $x$ in the context.

\subsection{Relational Semantics}

\subsubsection{Relation Environments}

For a set $X$, let $\Rel(X)$ be the set of binary relations on
$X$. For an index context $\Delta$, a \emph{relation environment} is a
family of functions $\rho = \{ \rho_{\tyPrimNm{X}} :
\idxTms{\Delta}{s} \to \Rel(\tyPrimSem{X}) \}_{s \in \mathit{Sort},
  \tyPrimNm{X} \in \mathit{PrimType}(s)}$.

We define our relational semantics of types with respect to a given
collection of relation environments: we assume that, for each index
context $\Delta$, we are given a set of relation environments
$\relEnv{E}(\Delta)$. This family must satisfy the following two
properties:
\begin{enumerate}
\item Closure under re-indexing by index context morphisms: if $\rho
  \in \relEnv{E}(\Delta)$ and $\sigma : \Delta' \Rightarrow \Delta$,
  then $\rho \circ \sigma^* \in \relEnv{E}(\Delta')$.
\item If we have a pair of relation environments $\rho_1 \in
  \relEnv{E}(\Delta', i\mathord:s')$ and $\rho_2 \in
  \relEnv{E}(\Delta)$, along with an index context morphism $\sigma :
  \Delta' \Rightarrow \Delta$, such that the outer edges of the
  following diagram commute, for all primitive types $\tyPrimNm{X} \in
  \mathit{PrimType}(s)$:
  \begin{displaymath}
    \xymatrix{
      {\idxTms{\Delta}{s}} \ar[r]^(.45){\pi^*_{i\mathord:s'}} \ar[d]_{\sigma^*}
      &
      {\idxTms{\Delta,i\mathord:s'}{s}} \ar[d]^{\sigma_{s'}^*} \ar@/^/[rdd]^{\rho_1}
      \\
      {\idxTms{\Delta'}{s}} \ar[r]^(.45){\pi^*_{i\mathord:s'}} \ar@/_/[rrd]_{\rho_2}
      &
      {\idxTms{\Delta',i\mathord:s'}{s}} \ar@{.>}[dr]_\rho
      \\
      &
      &
      {\Rel(X)}
    }
  \end{displaymath}
  Then there exists a relation environment $\rho \in
  \relEnv{E}(\Delta,i\mathord:s')$ such that the two triangles in the
  bottom right of the diagram commute.
\end{enumerate}

Given a relation environment $\rho \in \relEnv{E}(\Delta)$, we define
the set of extensions $\extends{\rho}{i\mathord:s}$ of $\rho$ by an
additional index variable $i\mathord:s$ to be the following:
\begin{displaymath}
  \extends{\rho}{i\mathord:s} = \{ \rho' \in \relEnv{E}(\Delta,i\mathord:s) \sepbar \rho' \circ \pi^*_{i\mathord:s} = \rho \}
\end{displaymath}

\subsubsection{Relational Semantics of Types}

%FIXME: re-word this
A given relation context $\rho \in \relEnv{E}(\Delta)$ is extended to
a relation interpretation of any type $\Delta \vdash A : \sortType$ by
induction its derivation. So we define a relation
$\rsem{A}{\relEnv{E}}{\rho} \in \Rel(\tySem{A})$ for each well-formed
type $A$.
\begin{eqnarray*}
  \rsem{\tyUnit}{\relEnv{E}}\rho & = & \{(*,*)\} \\
  \rsem{\tyPrim{X}{e}}{\relEnv{E}}\rho & = & \rho_{\tyPrimNm{X}}(e) \\
  \rsem{A \tyArr B}{\relEnv{E}}\rho & = & \rsem{A}{\relEnv{E}}\rho \relArrow \rsem{B}{\relEnv{E}}\rho \\
  \rsem{A \tyProduct B}{\relEnv{E}}\rho & = & \rsem{A}{\relEnv{E}}\rho \relTimes \rsem{B}{\relEnv{E}}\rho \\
  \rsem{A + B}{\relEnv{E}}\rho & = & \rsem{A}{\relEnv{E}}\rho \relSum \rsem{B}{\relEnv{E}}\rho \\
  \rsem{\forall i\mathord:s.A}{\relEnv{E}}\rho & = & \bigcap\{ \rsem{A}{\relEnv{E}}\rho' \sepbar \rho' \in \extends{\rho}{i:s} \}
\end{eqnarray*}
In this definition we have made use of the following three
constructions on binary relations. If $R \in \Rel(X)$ and $S \in
\Rel(Y)$, then $R \relArrow S \in \Rel(X \to Y)$ is defined as $\{
(f_1,f_2) \sepbar \forall (a_1,a_2) \in R.\ (f_1a_1,f_2a_2) \in S
\}$. With the same assumptions on $R$ and $S$, the relation $R
\relTimes S \in \Rel(X \times Y)$ is defined as $\{
((a_1,b_1),(a_2,b_2)) \sepbar (a_1,a_2) \in R \land (b_1,b_2) \in S
\}$. Finally, the relation $R \relSum S \in \Rel(X + Y)$ is defined as
$\{ (\mathrm{inl}\ x, \mathrm{inl}\ x') \sepbar (x,x') \in R \} \cup
\{ (\mathrm{inr}\ y, \mathrm{inr}\ y') \sepbar (y,y') \in S \}$.

We extend the relational semantics to contexts of well-formed types by
the following clauses:
\begin{eqnarray*}
  \rsem{\epsilon}{\relEnv{E}}{\rho} & = & \id_{\ctxtSem{\epsilon}} \\
  \rsem{\Gamma, x : A}{\relEnv{E}}{\rho} & = & \rsem{\Gamma}{\relEnv{E}}\rho \relTimes \rsem{A}{\relEnv{E}}\rho
\end{eqnarray*}

\subsubsection{Parametricity}

This relies on \lemref{lem:tyeq-erasure} to make sure that it is
well-typed.

\begin{lemma}\label{lem:tyeq-rel}
  If $\Delta \vdash A \equiv B : \sortType$, then for all $\rho \in
  \relEnv{E}(\Delta)$, $\rsem{A}{\relEnv{E}}\rho = \rsem{B}{\relEnv{E}}\rho$.
\end{lemma}

This next lemma relies on \lemref{lem:tysubst-erasure} to make sure
that it is well-typed.

\begin{lemma}\label{lem:tysubst-rel}
  If $\Delta \vdash A : \sortType$ and $\sigma : \Delta' \Rightarrow
  \Delta$, then for all $\rho \in \relEnv{E}(\Delta')$,
  $\rsem{\sigma^*A}{\relEnv{E}}\rho = \rsem{A}{\relEnv{E}}{(\rho \circ \sigma^*)}$. Likewise,
  if $\Delta \vdash \Gamma \isCtxt$ and $\sigma : \Delta' \Rightarrow
  \Delta$, then for all $\rho \in \relEnv{E}(\Delta')$,
  $\rsem{\sigma^*\Gamma}{\relEnv{E}}\rho = \rsem{\Gamma}{\relEnv{E}}{(\rho \circ \sigma^*)}$.
\end{lemma}

>>>>

@@

Data:
\begin{enumerate}
\item A set $\mathit{Sort}$ of sorts
\item A collection of primitive index operations $\mathit{IndexOp}$
\item A map $\indexOpArity : \mathit{IndexOp} \to \mathit{Sort}^* \times \mathit{Sort}$
\item A set $\mathit{PrimType}$ of primitive types
\item A map $\primTyArity : \mathit{PrimType} \to \mathit{Sort}^*$
\end{enumerate}

Then generate:
\begin{enumerate}
\item Well-sorted index expressions
\item Well-indexed types
\end{enumerate}

More data:
\begin{enumerate}
\item A set of primitive operations $\mathit{PrimOps}$
\end{enumerate}

This generates the well-typed terms.

For the semantics:
\begin{enumerate}
\item A ``relational'' meaning for index contexts. Can this be made
  separate? Make the point that it is non-compositional, which is
  useful. 
\item Underlying semantics for each type
\item Assume that the underlying semantics for types is not affected
  by the index environment (could it be?)
\item For each index context and primitive type, we demand a
  relational meaning for that type.
\item Relational meanings are closed under some operations.
\end{enumerate}

@@

\begin{figure*}[t]
  \centering
  Well-indexed Types

  \medskip
  Equality of well-indexed types
  \begin{mathpar}
    \inferrule* [right=TyEqPrim]
    {\primTyArity(\tyPrimNm{X}) = (s_1,...,s_n) \\
      \{\Delta \vdash e_i \equiv e'_i : s_i\}_{1\leq i \leq n}}
    {\Delta \vdash \tyPrim{X}{e_1,...,e_n} \equiv \tyPrim{X}{e'_1,..,e'_n} : \sortType}

@@

The rules for building the types of a system in our generic framework
are given in \autoref{fig:types}. The types of our system are built
from a set of primitive types $\mathit{PrimType}$.

@@
